---
title: "Instacart Customer Segmentation"
subtitle: "DATA 608 Spring 2025 - Story 5"
author: "Stephanie Chiang"
format: pdf
execute:
  echo: false
  output: false
---

This is a customer segmentation analysis for Instacart users based on:

-   the total number of orders made for the year

-   the total number of items purchased for the year

-   the average number of days between orders

-   the minimum number of days between orders

```{r lib}
#| message: false
#| warning: false
#| include: false

library(tidyverse)
library(broom)
library(ggrepel)
```

```{r import}
#| message: false
#| warning: false
#| include: false

users <- read.csv("DataStory6_files/user_features.csv")
orders <- read.csv("DataStory6_files/orders.csv")
```

```{r transform}
orders2 <- orders |>
  select(user_id, order_number, days_since_prior_order) |>
  group_by(user_id) |>
  summarise(
    total_orders = max(order_number),
    min_freq = min(days_since_prior_order, na.rm = TRUE),
    mean_freq = mean(days_since_prior_order, na.rm = TRUE))

users2 <- users |>
  select(user_id:frozen.juice) |>
  mutate(total_items = rowSums(across(packaged.cheese:frozen.juice))) |>
  select(user_id, total_items)

cust <- orders2 |> left_join(users2, join_by(user_id))
```

## Principal Component Analysis

```{r pca}
cust2 <- cust[c(2:5)] |> scale()
cust_pca <- cust2 |> prcomp(retx = TRUE)
cust_pca
```

```{r summary}
#| output: true

summary(cust_pca)
cust_pca$rotation
```

## Principal Component Analysis

-   The weights for total number of orders and mean frequency of days between orders on PC1 are about equal (though the other 2 variables are not far off).

-   For PC2, minimum number of days between orders has the highest absolute value.

-   PC1 and PC2 have a combined proportion of variance of about 0.9.

## PCA Biplot

Here are the projected data points, with the direction and strength of the original variables on the principal components shown as scaled arrows.

```{r scale-arrows}
loadings <- as.data.frame(cust_pca$rotation[, 1:2])
loadings$Variable <- rownames(loadings)
cust_pca_df <- as.data.frame(cust_pca$x)
scale_factor <- max(abs(cust_pca_df[, 1:2])) / max(abs(loadings[, 1:2]))
loadings_scaled <- loadings
loadings_scaled[, 1:2] <- loadings_scaled[, 1:2] * scale_factor * 0.8
```

```{r biplot}
#| output: true

cust_pca |>
  augment(cust2) |>
  ggplot(aes(.fittedPC1, .fittedPC2)) +
    geom_point(size = 0.5, color = "gray") +
    geom_segment(data = loadings_scaled, aes(x = 0, y = 0, xend = PC1, yend = PC2),
                 arrow = arrow(length = unit(0.2, "cm")), color = "navy") +
    geom_text_repel(data = loadings, aes(x = PC1, y = PC2, label = Variable),
                    size = 3, color = "navy") +
    theme_minimal() +
    labs(title = "PCA Biplot", x = "PC1", y = "PC2")
```

## PCA and K-Means Clusters

Here are the data points once again projected onto principal components, colored by their assigned clusters and marked at the centers.

```{r pca-k}
#| output: true
set.seed(123)

cust_pca_scores <- cust_pca_df[, 1:2]
kmeans_result <- kmeans(cust_pca_scores, centers = 3)
cust_pca_scores$Cluster <- as.factor(kmeans_result$cluster)

ggplot(cust_pca_scores, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 0.5) +
  geom_point(data = as.data.frame(kmeans_result$centers), 
             aes(x = PC1, y = PC2), 
             color = "black", shape = 4, size = 3, stroke = 1) +
  labs(title = "PCA + K-Means Clustering", x = "PC1", y = "PC2") +
  theme_minimal()
```

## K-Means Clusters

```{r km-cluster}
cust_km <- cust_pca |>
  augment(cust2) |>
  select(.fittedPC1:.fittedPC4) |>
  kmeans(centers = 3, nstart = 10)

cust_clustered <- cust |> 
  add_column(cluster = factor(cust_km$cluster))
head(cust_clustered)
```

The cluster assignments of each customer can also be plotted (based on total orders and minimum frequency, for example).

```{r km-plot}
#| output: true

ggplot(cust_clustered, aes(x = total_orders, y = min_freq)) + 
  geom_point(aes(color = cluster)) +
  labs(title = "Customer Clusters", x = "Total number of orders", y = "Minimum number of days between orders") +
  theme_minimal()
```

## Customer Segmentation Analysis

Though 3 was selected as a starting point for the number of clusters, this number can be adjusted to further examine customer data using PCA and K-Means clustering.
