---
title: "Do stricter gun laws reduce firearm gun deaths?"
subtitle: "DATA 608 Spring 2025 - Story 3"
author: "Stephanie Chiang"
format: revealjs
editor: visual
execute:
  echo: false
---

## Introduction

Gun violence is an important issue that uniquely affects the Unites States. This is an analysis of the relationship between gun laws, firearm ownership and gun death rates across all 50 states. Using data from various sources including the CDC, the associations between these numbers can easily be visualized.

## Data

We will be examining the following data:

-   total firearm mortality for each state per 100,000 persons

-   total number of firearm laws for each state

-   estimated household gun ownership for each state

```{r import}
library(tidyverse)

# Import firearm mortality data
fm_raw <- read.csv("firearm_mortality2022.csv")

# Import firearm laws data
laws_raw <- read.csv("firearm_laws.csv")

# Import gun ownership data
own_raw <- read.csv("gun-ownership-by-state.csv")
```

```{r clean}
# Clean the firearm dataframe
fm <- fm_raw[fm_raw$YEAR == 2022,] |>
  select(STATE, RATE) |>
  rename(DEATH_RATE = RATE)
fm$region <- tolower(state.name[match(fm$STATE, state.abb)])

# Clean the ownership dataframe
own <- own_raw |>
  select(state, GunOwnershipPercentage) |>
  rename(OWNERSHIP_RATE = GunOwnershipPercentage, region = state)
own$region <- tolower(own$region)

# Clean the laws dataframe
laws <- laws_raw |>
  select(STATE, LAWTOTAL) |>
  rename(region = STATE, LAW_TOTAL = LAWTOTAL)
laws$region <- tolower(laws$region)
```

```{r join}
# Join the dataframes
firearms <- merge(fm, own, by = "region", all = TRUE)
firearms <- merge(firearms, laws, by = "region", all = TRUE)
firearms <- firearms |> drop_na()
```

```{r data}
#| echo: true

head(firearms)
```

## Plotting the linear relationships

```{r mort-plot}
#| output-location: slide

ggplot(firearms, aes(x = DEATH_RATE, y = OWNERSHIP_RATE)) +
  geom_point() +
  labs(title = "Gun Ownership vs Mortality Rates for the Unites States",
       x = "Gun Ownership Rate",
       y = "Firearm Mortality Rate per 100,000 Persons") +
  theme_bw() + 
  geom_smooth(method = "lm", formula = "y ~ x")
```

---

```{r law-own-plot}
#| output-location: slide

ggplot(firearms, aes(x = OWNERSHIP_RATE, y = LAW_TOTAL)) +
  geom_point() +
  labs(title = "Gun Laws vs Ownership for the Unites States",
       x = "Total Number of Laws",
       y = "Gun Ownership Rate") +
  theme_bw() + 
  geom_smooth(method = "lm", formula = "y ~ x")
```

---

```{r laws-death}
#| output-location: slide
ggplot(firearms, aes(x = DEATH_RATE, y = LAW_TOTAL)) +
  geom_point() +
  labs(title = "Gun Ownership vs Mortality Rates for the Unites States",
       x = "Total Number of Laws",
       y = "Firearm Mortality Rate per 100,000 Persons") +
  theme_bw() + 
  geom_smooth(method = "lm", formula = "y ~ x")
```

## Mapping the data

```{r joins}
library(maps)

cont_states <- map_data("state")
cont_states <- cont_states |> select(!subregion)

alaska <- map_data("world") |> subset(subregion == "Alaska")
alaska$region <- "alaska"
alaska <- alaska |> select(!subregion)

hawaii <- map_data("world") |> subset(subregion == "Hawaii")
hawaii$region <- "hawaii"
hawaii <- hawaii |> select(!subregion)

all_states <- merge(cont_states, alaska, all = TRUE)
all_states <- merge(all_states, hawaii, all = TRUE)

cont_guns <- left_join(cont_states, firearms, by = "region")
all_guns <- left_join(all_states, firearms, by = "region") |>
  rename(abb = STATE, state = region)
```

```{r map-mort}
#| message: true
#| warning: false
#| output-location: slide

library(usmap)
library(patchwork)

map1 <- plot_usmap(data = all_guns,
           values = "DEATH_RATE",
           regions = "states",
           color = "gray90",
           size = 0.1) +
  scale_fill_continuous(low = "#ADD8E6", high = "#132B43", name = "Death Rate") + 
  labs(title = "Firearm Mortality Rate per 100,000 Persons") +
  theme(legend.position = "right")

map2 <- plot_usmap(data = all_guns,
           values = "LAW_TOTAL",
           regions = "states",
           color = "gray90",
           size = 0.1) +
  scale_fill_continuous(low = "#132B43", high = "#ADD8E6", name = "Number of Laws") + 
  labs(title = "Presence of Gun Laws") +
  theme(legend.position = "right")

map1 + map2
```

## Conclusion

These plots and maps suggest clear links between gun laws, gun ownership and gun deaths. States with more laws on the books tend to also have lower gun ownership rates and states with high gun ownership tend to report higher mortality rates.
